Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-02-03 paper/zhang-cook-laa-langrene-menendez.Rmd
minor fixes

Apart from checking the search and progression of an optimiser, looking at where the bases are positioned in the basis space is another interest. Given the orthonormality constraint, the space of projection bases $\mathbf{A}_{p \times d}$ is a $p \times d$ dimension sphere and  dimension reduction methods, i.e. principal component analysis is applied to first project all the bases onto a 2D space. In a projection pursuit guided tour optimisation, there are various type of bases involved: 1) The start basis; 2) The search bases that an optimiser evaluated to produce the anchor bases; 3) The anchor bases that have the highest index value in each iteration; 4) The interpolating bases on the interpolation path; and finally 5) the end basis. The importance of these basis differs and the most important ones are the start, interpolating and end bases. Anchor and search bases can be turned on with argument `details = TRUE`. Sometimes, two optimisers can start with the same basis but finish with bases of opposite sign. This happens because the projection is invariant to the sign difference of the bases and so does the index value, however, this creates difficulties for comparing the optimisers since the end bases will be symmetric with respect to the origin. A sign flipping step is conducted to flip the signs of all the bases in one routine if different optimisations finish at opposite places.

Several annotations have been made to help understanding this plot. As mentioned previously, the original basis space is a high dimensional sphere and random bases on the sphere can be generated via the CRAN package \CRANpkg{geozoo} [@geozoo]. Along with the bases recorded during the optimisation and a zero basis, the parameters (centre and radius) of the 2D space can be estimated. PCA is performed to get the first two PC coordinates of all the bases and the centre of the circle is the PC coordinates of the zero matrix, and radius is estimated as the largest distance from the centre to the basis. The theoretical best basis is known for simulated data and can be labelled to compare the how close the end basis found by the optimisers is to the theoretical best. Various aesthetics, i.e. size, alpha and colour, are applicable to emphasize critical elements and adjust for the presentation. For example, anchor points and search points are less important and hence a smaller size and alpha is used; Alpha can also be applied on the interpolation paths to show the start to finish from transparent to opaque.

@@

Animation is another display to show how the search progresses from start to finish in the space and an  `animate = TRUE` argument is used to enable the animation in the PCA plot. Figure \ref{fig:toy-pca-animated} shows six frames from the animation of the PCA plot in Figure \ref{fig:toy-pca}. An additional piece of information one can learn from this animation is that SA finds its end basis quicker than PD since SA finishes its search in the 5th frame while PD is still making more progression.

@@

```{r toy-tour, fig.cap="Six selected frames from the tour animation for viewing the same two optmisations as previous two figures. The tour animation allows for an appreciation of the search bases in the original high dimensional space. A full video of this aimation can be found at \\url{https://vimeo.com/user132007430/review/504328122/9be84db563}"}
```

As mentioned previously, the original $p \times d$ dimension space can be simulated via randomly generated bases in \CRANpkg{geozoo} [@geozoo] package. While the PCA plot projects the bases from the direction that maximises the variance, the tour plot display of the original high dimensional space from various directions using animation. Figure \ref{fig:toy-tour} shows frames from the tour plot of the same two optimisations in its original 5D space.

@@

@@

Random variables with different distributions have been simulated and the distributional form of each variable is presented in Equations \ref{eq:sim-norm} to \ref{eq:sim-x7}. Variable `x1`, `x8`, `x9` and `x10` are normal noise with zero mean and unit variance and `x2` to `x7` are normal mixtures with varied weights and locations. All the variables have been scaled to have an overall unit variance before projection pursuit.

@@

An example of non-monotonic interpolation has been given in Figure \ref{fig:toy-interp}: a path that passes bases with higher index value than the target one. For SAJO, a non-monotonic interpolation is justified since target bases do not necessarily have a higher index value than the current one, while this is not the case for SA. The original trace plot for a 2D projection problem, optimised by SA, is shown on the left panel of Figure \ref{fig:interruption} and one can observe clearly that the non-monotonic interpolation has undermined the optimiser to realise its full potential. Hence, an interruption is implemented to stop at the best basis found in the interpolation. The right panel of Figure \ref{fig:interruption} shows the trace plot after implementing the interruption and while the first two interpolations are identical, the basis at Time 61 has a higher index value than the target in the third interpolations. Rather than starting the next iteration from the target basis on Time 65, SA starts the next iteration at Time 61 on the right panel and reaches a better final basis.

@@

```{r noisy-better-geo, fig.height = 4, fig.cap = "Trace and PCA plot of 1D projection on 5\\-variable dataset, \\code{boa5}, with \\code{norm\\_kol} index, optimised by all three optimiers. Optimiser PD fails to optimise the non-smooth index while although both SA and SAJO success in optimise the index, SAJO takes much longer than SA and finishes off closer to the theoretial best."}
```

Figure \ref{fig:noisy-better-geo} presents the trace and PCA plots of all three optimisers and as expected, none of the interpolated path is smooth. There is barely any improvement made by PD, indicating its failure in optimising non-smooth index. While SA and SAJO have managed to get close to the index value of the theoretical best, trace plot shows that it takes SAJO much longer to interpolate towards the final basis. This long interpolation path is partially due to the fluctuation in the early iterations, where the SAJO tends to generously accept inferior bases before concentrating near the optimum. The PCA plot shows the interpolation path and search points excluding the last termination iteration. [A sentence for PD]. Comparing the amount of random search done by SA and SAJO, it is surprising that SAJO doesn't carry that many of sampling as SA. Combine the insight from both the trace and PCA plot, one can learn the two different search strategies by SA and SAJO: SA frequently samples in the space and only make a move when an improvement is guaranteed to be made, while SAJO first broadly accepts bases in the space and then starts the extensive sampling in a narrowed subspace. The specific decision of which optimiser to use will depend on the index curve in the basis space but if the basis space is non-smooth, accepting inferior bases at first, as what SAJO has done, can lead to a more efficient search, in terms of the overall number of points evaluated.

```{r kol-result, fig.height = 4, fig.cap="PCA plot of 1D projection on the 6\\-variable data, \\code{boa6}, with normal kolmogorov index. Two optimisers: SA and SAJO and two search sizes: 0.5 and 0.7 are used in the simulation for 20 replicates. Local optimum (x), is presented in this experiment, along with the global optimum (*). The final basis is larger and more opaque than the start basis for each path and color represents whether the global or local optimum is found, after polishing."}
```

The next experiment compares the performance of SA and SAJO when a local maximum exists. Two search neighbourhood sizes: 0.5 and 0.7 are compared to understand how a large search neighbourhood would affect the final basis and the length of the search. Figure \ref{fig:kol-result} shows 80 paths simulated using 20 seeds in the PCA plot, faceted by  optimiser and search size. With SA and a search size of 0.5, despite being the simplest and fastest, the optimiser fails in three instances where the final basis lands neither near the local nor the global optimum. With a larger search size of 0.7, more seeds have found the global maximum. Comparing SA and SAJO for a search size of 0.5, SAJO doesn't seem to improve the final basis found, despite having longer interpolation paths. However, the denser paths near the local maximum is an indicator that SAJO is working hard to examine if there is any other optimum in the basis space but the relative small search size has diminished its ability to reach the global maximum. With a larger search size, almost all the seeds (16 out of 20) have found the global maximum and some final bases are much closer to the theoretical best, as compared to the three other cases. This indicates that SAJO, with a reasonable large search size, is able to overcome the local optimum and optimise close towards the global optimum.

@@

- Utilities
  - \code{theme\_fern()} and \code{format\_label()} for better display of the grid lines and axis formatting
  - \code{clean\_method()} for clean up the name of the optimisers
  - \code{botanical\_palettes} is a collection of colour palettes from Australian native plants Quantitative palettes include daisy, banksia and cherry and sequential palettes contain fern and acacia
  - \code{botanical\_pal()} as the colour interpolator 
  - \code{scale\_color\_*()} and \code{scale\_fill\_*()} for scaling the colour and fill of the plot

@@

<!-- summarise what we have done and what might be done in the future. -->
Most of the work in this project has been translated into code in two packages: the collection of the data object is implemented in the existing \CRANpkg{tourr}[@tourr] package; manipulation and visualisation of the data object are implemented in the new \pkg{ferrn} package. Equipped with handy tools to diagnose the performance of optimisers, future work can extend the diagnostics to a wider range of index functions, i.e. scagnostics, association, and information index [@laa2020using] and understand how the optimisers behave for index functions with different structures.

@@

This article is created using \CRANpkg{knitr}[@knitr] and \CRANpkg{rmarkdown} [@rmarkdown]  in R. The source code for reproducing this paper can be found at: \url{https://github.com/huizezhang-sherry/paper-tour-vis}.

>>>>

Apart from checking the search and progression of an optimiser, looking at where the bases are positioned in the basis space is another interest. Given the orthonormality constraint, the space of projection bases $\mathbf{A}_{p \times d}$ is a $p \times d$-dimensional sphere. A dimensionality reduction method, e.g. principal component analysis is applied to first project all the bases onto a 2D space. In a projection pursuit guided tour optimisation, there are various types of bases involved: 1) The starting basis; 2) The search bases that the optimiser evaluated to produce the anchor bases; 3) The anchor bases that have the highest index value in each iteration; 4) The interpolating bases on the interpolation path; and finally 5) the end basis. The importance of these bases differs and the most important ones are the starting, interpolating and end bases. The anchor and search bases can be turned on with argument `details = TRUE`. Sometimes, two optimisers can start with the same basis but finish with bases of opposite sign. This happens because the projection is invariant to the sign difference of the bases and so does the index value, however, this creates difficulties for comparing the optimisers since the end bases will be symmetric with respect to the origin. A sign flipping step is conducted to flip the signs of all the bases in one routine if different optimisations finish at opposite places.

Several annotations have been made to help understanding this plot. As mentioned previously, the original basis space is a high-dimensional sphere and random bases on the sphere can be generated via the CRAN package \CRANpkg{geozoo} [@geozoo]. Along with the bases recorded during the optimisation and a zero basis, the parameters (centre and radius) of the 2D space can be estimated. PCA is performed to get the first two PC coordinates of all the bases and the centre of the circle is the PC coordinates of the zero matrix, and radius is estimated as the largest distance from the centre to the basis. The theoretical best basis is known for simulated data and can be labelled to compare how close the end basis found by the optimisers is to the theoretical best. Various aesthetics, i.e. size, alpha and colour, are applicable to emphasize critical elements and adjust for the presentation. For example, anchor points and search points are less important and hence a smaller size and alpha is used. Alpha can also be applied on the interpolation paths to show the start to finish from transparent to opaque.

@@

Animation is another type of display to show how the search progresses from start to finish in the space. An  `animate = TRUE` argument is used to enable the animation in the PCA plot. Figure \ref{fig:toy-pca-animated} shows six frames from the animation of the PCA plot in Figure \ref{fig:toy-pca}. An additional piece of information one can learn from this animation is that SA finds its end basis quicker than PD since SA finishes its search in the 5th frame while PD is still making more progress.

@@

```{r toy-tour, fig.cap="Six selected frames from the tour animation for viewing the same two optimisations as previous two figures. The tour animation allows for an appreciation of the search bases in the original high dimensional space. A full video of this aimation can be found at \\url{https://vimeo.com/user132007430/review/504328122/9be84db563}"}
```

As mentioned previously, the original $p \times d$ dimension space can be simulated via randomly generated bases in the \CRANpkg{geozoo} [@geozoo] package. While the PCA plot projects the bases from the direction that maximises the variance, the tour plot displays the original high-dimensional space from various directions using animation. Figure \ref{fig:toy-tour} shows some frames from the tour plot of the same two optimisations in its original 5D space.

@@

\textcolor{red}{Something to fill here ...}

@@

Random variables with different distributions have been simulated and the distributional form of each variable is presented in equations \ref{eq:sim-norm} to \ref{eq:sim-x7}. Variables `x1`, `x8`, `x9` and `x10` are normal noise with zero mean and unit variance and `x2` to `x7` are normal mixtures with varied weights and locations. All the variables have been scaled to have an overall unit variance before projection pursuit.

@@

An example of non-monotonic interpolation has been given in Figure \ref{fig:toy-interp}: a path that passes bases with higher index value than the target one. For SAJO, a non-monotonic interpolation is justified since target bases do not necessarily have a higher index value than the current one, while this is not the case for SA. The original trace plot for a 2D projection problem, optimised by SA, is shown on the left panel of Figure \ref{fig:interruption} and one can observe clearly that the non-monotonic interpolation has undermined the optimiser to realise its full potential. Hence, an interruption is implemented to stop at the best basis found in the interpolation. The right panel of Figure \ref{fig:interruption} shows the trace plot after implementing the interruption and while the first two interpolations are identical, the basis at Time 61 has a higher index value than the target in the third interpolation. Rather than starting the next iteration from the target basis on Time 65, SA starts the next iteration at Time 61 on the right panel and reaches a better final basis.

@@

```{r noisy-better-geo, fig.height = 4, fig.cap = "Trace and PCA plot of 1D projection on 5\\-variable dataset, \\code{boa5}, with \\code{norm\\_kol} index, optimised by all three optimisers. Optimiser PD fails to optimise the non-smooth index while although both SA and SAJO succeed in optimising the index, SAJO takes much longer than SA and finishes off closer to the theoretial best."}
```

Figure \ref{fig:noisy-better-geo} presents the trace and PCA plots of all three optimisers and as expected, none of the interpolated path is smooth. There is barely any improvement made by PD, indicating its failure in optimising non-smooth index. While SA and SAJO have managed to get close to the index value of the theoretical best, trace plot shows that it takes SAJO much longer to interpolate towards the final basis. This long interpolation path is partially due to the fluctuation in the early iterations, where the SAJO tends to generously accept inferior bases before concentrating near the optimum. The PCA plot shows the interpolation path and search points excluding the last termination iteration. \textcolor{red}{[A sentence for PD]}. Comparing the amount of random search done by SA and SAJO, it is surprising that SAJO does not carry as many samples as SA. Combining the insights from both the trace and PCA plot, one can learn the two different search strategies by SA and SAJO: SA frequently samples in the space and only make a move when an improvement is guaranteed to be made, while SAJO first broadly accepts bases in the space and then starts the extensive sampling in a narrowed subspace. The specific decision of which optimiser to use will depend on the index curve in the basis space but if the basis space is non-smooth, accepting inferior bases at first, as what SAJO has done, can lead to a more efficient search, in terms of the overall number of points evaluated.

```{r kol-result, fig.height = 4, fig.cap="PCA plot of 1D projection on the 6\\-variable data, \\code{boa6}, with normal kolmogorov index. Two optimisers: SA and SAJO and two search sizes: 0.5 and 0.7 are used in the simulation for 20 replicates. The local optimum (x), is presented in this experiment, along with the global optimum (*). The final basis is larger and more opaque than the start basis for each path and color represents whether the global or local optimum is found, after polishing."}
```

The next experiment compares the performance of SA and SAJO when a local maximum exists. Two search neighbourhood sizes: 0.5 and 0.7 are compared to understand how a large search neighbourhood would affect the final basis and the length of the search. Figure \ref{fig:kol-result} shows 80 paths simulated using 20 seeds in the PCA plot, faceted by  optimiser and search size. With SA and a search size of 0.5, despite being the simplest and fastest, the optimiser fails in three instances where the final basis lands neither near the local nor the global optimum. With a larger search size of 0.7, more seeds have found the global maximum. Comparing SA and SAJO for a search size of 0.5, SAJO does not seem to improve the final basis found, despite having longer interpolation paths. However, the denser paths near the local maximum is an indicator that SAJO is working hard to examine if there is any other optimum in the basis space but the relative small search size has diminished its ability to reach the global maximum. With a larger search size, almost all the seeds (16 out of 20) have found the global maximum and some final bases are much closer to the theoretical best, as compared to the three other cases. This indicates that SAJO, with a reasonable large search size, is able to overcome the local optimum and optimise close towards the global optimum.

@@

- Utilities
  - \code{theme\_fern()} and \code{format\_label()} for better display of the grid lines and axis formatting
  - \code{clean\_method()} to clean up the name of the optimisers
  - \code{botanical\_palettes} is a collection of colour palettes from Australian native plants. Quantitative palettes include daisy, banksia and cherry and sequential palettes contain fern and acacia
  - \code{botanical\_pal()} as the colour interpolator 
  - \code{scale\_color\_*()} and \code{scale\_fill\_*()} for scaling the colour and fill of the plot

@@

<!-- summarise what we have done and what might be done in the future. -->
Most of the work in this project has been translated into code in two packages: the collection of the data object is implemented in the existing \CRANpkg{tourr} [@tourr] package; while the manipulation and visualisation of the data object are implemented in the new \pkg{ferrn} package. Equipped with handy tools to diagnose the performance of optimisers, future work can extend the diagnostics to a wider range of index functions, i.e. scagnostics, association, and information index [@laa2020using] and understand how the optimisers behave for index functions with different structures.

@@

This article is created using \CRANpkg{knitr} [@knitr] and \CRANpkg{rmarkdown} [@rmarkdown]  in R. The source code for reproducing this paper can be found at: \url{https://github.com/huizezhang-sherry/paper-tour-vis}.
